Final analysis of Greenland narwhals trajectories with smoothSDE

### 2025-01-19 R/application/CRCVM/fit_crcvm.Rmd
try fit with DistanceShore_translated and theta_translated

#Fixed measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre12),dim=c(2,2,n_pre12))

#parameters formulas
formulas <- list(tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~s(theta,k=6,bs="cs"),D0=~1)

# Fit baseline
baseline_crcvm_cubic<- SDE$new(formulas = formulas,data = renamed_dataBE12,type = "CRCVM_SSM",
                    response = c("x","y"),par0 = c(2,4,0,0.05),other_data=list("H"=H),
                    fixpar=c("D0"))

baseline_crcvm_cubic$update_lambda(c(exp(3),exp(3),10^-12))
baseline_crcvm_cubic$update_map(list("log_lambda"=factor(c(1,2,NA))))
baseline_crcvm_cubic$fit()

>>>>

#Fixed measurement error
sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_pre12),dim=c(2,2,n_pre12))

#parameters formulas
formulas <- list(tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                   a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)

# Fit baseline
baseline_crcvm_cubic<- SDE$new(formulas = formulas,data = renamed_dataBE12,type = "CRCVM_SSM",
                    response = c("x","y"),par0 = c(1,4,0.1,4,0.2,0.5,0.1,pi/6),other_data=list("H"=H),
                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"))

baseline_crcvm_cubic$fit()

### 2025-01-19 R/application/RCVM/fit_rcvm.Rmd
try fit with DistanceShore_translated and theta_translated

#data before exposure with first 24h removed
dataBE24=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")

#data before exposure with first 12h removed
dataBE12=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")

@@

D_low=0.073
D_up=3
dataBE24[dataBE24$DistanceShore> D_up,"Ishore"]=0
dataBE24[dataBE24$DistanceShore< D_low,"Ishore"]=1/D_low

@@

@@

The first model is a CTCRW where the covariate AngleNormal influences the persistence of the movement.

@@

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"))

@@

#Fit baseline with measurement error estimated from the data
baseline_ctcrw0<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
                          response = c("x","y"),
                          par0 = par0,
                          other_data=list("H"=H),
                          fixpar=c("mu1","mu2"))
baseline_ctcrw0$fit()

@@

```{r plot smooth parameters baseline_ctcrw0 }
baseline_ctcrw0$plot_par(var="AngleNormal",
                         par_names=c("tau","nu"),n_post=100)

@@

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                 omega=~s(AngleNormal,k=10,bs="cs"))

@@

# Fit baseline with measurement error estimated from the data
baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))

@@

```{r plot smooth parameters baseline_racvm0}
baseline_racvm0$plot_par(var="AngleNormal",
                         par_names=c("omega","tau"),n_post=100)

@@

## Baseline with AngleNormal and Ishore in omega and tau

@@

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(Ishore,k=5,bs="cs")+
                   ti(AngleNormal,Ishore,k=c(5,5),bs="cs"))

# Fit baseline
baseline_racvm1_unstable<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"))

@@

@@

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=exp(c(3,3,4,-2,-2,-2,-2))

#mapping for fixed smoothing penalties
new_map=list("log_lambda"=factor(c(1,2,3,rep(NA,4))))

@@

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(Ishore,k=5,bs="cs")+
                   ti(AngleNormal,Ishore,k=c(5,5),bs="cs"))

# Fit baseline
baseline_racvm1<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"))

@@

@@

xmin=list("Ishore"=1/D_up)
xmax=list("Ishore"=1/D_low)
link=list("Ishore"=(\(x) 1/x))
xlabel=list("Ishore"="Distance to shore")

plots_bas_racvm1=baseline_racvm1$get_all_plots(baseline=NULL,
      xmin=xmin, xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

@@

#initial parameters
par0 <- c(0,0,2,4,0)
init_lambda=c(exp(4),exp(4),10^-6,10^-1)

@@

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(AngleNormal,DistanceShore,k=5,bs="cs"))

@@

xmin=list("Ishore"=1/D_up)
xmax=list("Ishore"=1/D_low)
link=list("Ishore"=(\(x) 1/x))
xlabel=list("Ishore"="Distance to shore")

plots_bas_racvm2=baseline_racvm2$get_all_plots(baseline=NULL,
      xmin=xmin, xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

@@

@@

The adjusted SDE models where the smooth parameters $\omega$ has been multiplied by a factor $40$ produces trajectories that seem to follow the shoreline and look more similar to the actual data.
However, the trajectories obtained from the non adjusted fitted model does not produce constrained trajectories.
This suggests that we are indeed underestimating the smooth parameter $\omega$ with the tensor splines. This may be due to several factors : the low frequency at which the data is observed, the measurement error, the small amount of data that we have.

We can try to fit an interpretable parametric surface on the estimated spline surface for the parameter $\omega$ in the models \texttt{baseline\_racvm1} or \texttt{baseline\_racvm2}. However the surfaces are so different that it does not work very well.

Ishore=seq(from=1/6,to=1/0.05,length.out=30)
theta=seq(from=-pi,to=pi,length.out=30)
grid<- as.data.frame(expand.grid(Ishore,theta))
colnames(grid) <- c("Ishore","AngleNormal")
grid$ID="Asgeir"

#get model matrices
mats=baseline_racvm1$make_mat(new_data=grid)
X_fe=mats$X_fe
X_re=mats$X_re

par_mat=baseline_racvm1$par(new_data=grid,X_fe=X_fe,X_re=X_re)

# Define the parametric function for omega
fomega<- function(cov_data, a =5,theta0=pi/2,D0=1) {
  Dshore=cov_data$DistanceShore
  theta=cov_data$AngleNormal
  if (is.null(Dshore)){
    Dshore=1/cov_data$Ishore
  }
  a*theta*(theta-theta0)*(theta+theta0)*exp(-Dshore/D0)/Dshore
}

#fix magnitude and slope for omega on the AngleNormal axis
a_fixed <- 1  
theta0_fixed <- pi/2

fit_optim <- optim(par = c(log_D0 = log(1)),  # Exclude a from parameters
                   fn = function(params) {
                     log_D0 <- params[1]

                     # Use fixed values
                     a <- a_fixed
                     theta0 <- theta0_fixed

                      # Compute predicted z values
                     z_pred <- fomega(data, a,theta0,exp(log_D0))

# Compute the predicted values using the starting parameters
z_pred_starting <- fomega(grid,D0=exp(est_par))

# Reshape z_pred_starting to a matrix for plotting
z_pred_matrix <- matrix(z_pred_starting, nrow = length(Ishore), ncol = length(theta))

# Create the 3D surface plot using plot_ly
plot_ly(x = ~theta, y = ~1/Ishore, z = ~z_pred_matrix) %>%
  add_surface() %>%
  layout(title = "Parametric surface for omega (starting parameters)",
         scene = list(
           yaxis = list(title = "1/Ishore"),
           xaxis = list(title = "AngleNormal"),
           zaxis = list(title = "Predicted z")
         ))

# Create the 3D surface plot using plot_ly
plot_ly(x = ~theta, y = ~1/Ishore, z = ~z) %>%
  add_surface() %>%
  layout(title = "Predicted spline surface for omega",
         scene = list(
           yaxis = list(title = "Ishore"),
           xaxis = list(title = "AngleNormal"),
         ))

@@

## Try a model similar to baseline racvm2

@@

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                 omega=~te(AngleNormal,Ishore,k=5,bs="cs"))

response0<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM_SSM",response = c("x","y"),
                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))

response0$update_map(list("log_lambda"=factor(c(1,2,NA,NA))))
response0$update_lambda(exp(c(4,4,-5,-5)))

response0$fit()
```
```{r}
response0$get_all_plots(baseline=NULL,
      xmin=xmin, xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
```

>>>>

#data before exposure with first 24h removed
dataBE24=read.csv(file.path(narwhal_data_path,"DataBE24.csv"), header = TRUE,dec = ".")

#data before exposure with first 12h removed
dataBE12=read.csv(file.path(narwhal_data_path,"DataBE12.csv"), header = TRUE,dec = ".")

@@

D_low=0.05
D_up=3
dataBE24[dataBE24$DistanceShore> D_up,"Ishore"]=0
dataBE24[dataBE24$DistanceShore< D_low,"Ishore"]=1/D_low

@@

all_data <- data.frame()

for (id in unique(dataBE12$ID)) {

  sub_dataBE <- dataBE12[dataBE12$ID == id, ]
  sub_dataAE <- dataAE[dataAE$ID == id, ]

  sub_data <- rbind(sub_dataBE, sub_dataAE)

  all_data <- rbind(all_data, sub_data)
}

@@

The first model is a CTCRW where the covariate theta influences the persistence of the movement.

@@

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"))

@@

#Fit baseline with measurement error estimated from the data
baseline_ctcrw0<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
                          response = c("x","y"),
                          par0 = par0,
                          other_data=list("log_sigma_obs0"=log(0.05)),
                          fixpar=c("mu1","mu2"))
baseline_ctcrw0$fit()

@@

```{r plot smooth parameters baseline_ctcrw0 }
baseline_ctcrw0$plot_par(var="theta_translated",
                         par_names=c("tau","nu"),n_post=100)

@@

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                 omega=~s(theta_translated,k=10,bs="cs"))

@@

# Fit baseline with measurement error estimated from the data
baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",response = c("x","y"),
                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))

@@

```{r plot smooth parameters baseline_racvm0}
baseline_racvm0$plot_par(var="theta_translated",
                         par_names=c("omega","tau"),n_post=100)

@@

## Baseline with AngleNormal and DistanceShore in omega and tau

@@

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,DistanceShore_translated,k=5,bs="cs"))

# Fit baseline
baseline_racvm1_unstable<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"))

@@

First, we get good values for the smoothing penalty by approximating the true surface with splines.

fomega=function(cov_data,a=2,b=5,D0=0.5,D1=0.8,sigma_D=0.3,sigma_theta=pi/6){
  Dshore=cov_data$DistanceShore
  theta=cov_data$theta
  if (is.null(Dshore)){
    Dshore=1/cov_data$Ishore
  }

  a*theta*(theta-pi/2)*(theta+pi/2)*exp(-Dshore/D0)/Dshore+
      b*(exp(-1/2*(((theta+pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2))-
           exp(-1/2*(((theta-pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2)))
}

# Splines smooth for omega -------------------- 
n <- 100000
SP_DF=5

#sample theta and DistanceShore points in the domain
theta <- runif(n,-pi,pi);
DistanceShore <- runif(n,0.02,7)
samples=data.frame(theta=theta,DistanceShore=DistanceShore)

#define grid of values
theta_v <- seq(-pi,pi,length=30)
Dshore_v<- seq(0.02,7,length=30)
pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))

#points on the surface perturbed by gaussian noise
f <- fomega(samples)
y <- f+0.1*rnorm(n)

#fit with bivariate splines te
m1 <- gam(y~te(theta,DistanceShore,k=SP_DF))

@@

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(exp(3),exp(3),1,exp(-12),exp(-5))

#mapping for fixed smoothing penalties
new_map=list("log_lambda"=factor(c(1,2,3,rep(NA,2))))

@@

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,DistanceShore_translated,k=c(5,5),bs="cs"))

# Fit baseline
baseline_racvm1<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
                    response = c("x","y"),par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
                    fixpar=c("mu1","mu2"))

@@

```{r results baseline_racvm1}

estimates_bas_racvm1=as.list(baseline_racvm1$tmb_rep(),what="Est")
std_bas_racvm1=as.list(baseline_racvm1$tmb_rep(),what="Std")
```

@@

xmin=list("DistanceShore_translated"=0.2)
xmax=list("DistanceShore_translated"=2)

plots_bas_racvm1=baseline_racvm1$get_all_plots(baseline=NULL,
      xmin=xmin, xmax=xmax,show_CI="pointwise")

@@

#initial parameters
par0 <- c(0,0,2,4,0)
init_lambda=c(exp(3),exp(3),exp(-12),exp(-5))

@@

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,DistanceShore_translated,k=5,bs="cs"))

@@

xmin=list("DistanceShore_translated"=0.2)
xmax=list("DistanceShore_translated"=4)

plots_bas_racvm2=baseline_racvm2$get_all_plots(baseline=NULL,
      xmin=xmin,xmax=xmax,show_CI="pointwise")

@@

estimate_parametric_omega=function(model) {

    data=model$data()

    Dshore=seq(from=quantile(data$DistanceShore_translated,0.5),to=quantile(data$DistanceShore_translated,0.95),length.out=30)
    theta_translated=seq(from=-pi,to=pi,length.out=30)
    grid<- as.data.frame(expand.grid(Dshore,theta_translated))
    colnames(grid) <- c("DistanceShore_translated","theta_translated")
    grid$ID="Asgeir"

    #get model matrices
    mats=model$make_mat(new_data=grid)
    X_fe=mats$X_fe
    X_re=mats$X_re

    par_mat=model$par(new_data=grid,X_fe=X_fe,X_re=X_re)

    # Create the 3D surface plot using plot_ly
    plot_splines<- plot_ly(x = ~theta_translated, y = ~Dshore, z = ~z) %>%
      add_surface() %>%
      layout(title = "Predicted spline surface for omega",
             scene = list(yaxis = list(title = "Distance to shore"),
               xaxis = list(title = "theta_translated")))

    fit_optim <- optim(par = c(0.1,2),
                       fn = function(params) {

                        # Compute predicted z values
                        z_pred <- fomega(data,1,0.5,params[1],params[2],1,pi/6)

    print(est_par)

    # Compute the predicted values using the starting parameters
    z_pred_starting <- fomega(grid,1,1,est_par[1],est_par[2],1,
                                    pi/6)

    # Reshape z_pred_starting to a matrix for plotting
    z_pred_matrix <- matrix(z_pred_starting, nrow = length(Dshore), ncol = length(theta_translated))

    print(z_pred_matrix)

    # Create the 3D surface plot using plot_ly
    plot_parametric <- plot_ly(x = ~theta_translated, y = ~Dshore, z = ~z_pred_matrix) %>%
      add_surface() %>%
      layout(title = "Estimated parametric surface for omega",
             scene = list(
               yaxis = list(title = "Distance to shore"),
               xaxis = list(title = "theta"),
               zaxis = list(title = "Predicted z")
             ))

    return (list("est_par"=est_par,"splines_surface"=plot_splines,"parametric_surface"=plot_parametric))

}

estimate_parametric_omega(baseline_racvm2)

@@

The adjusted SDE models where the smooth parameters $\omega$ has been multiplied by a factor $40$ produces trajectories that seem to follow the shoreline and look more similar to the actual data.
However, the trajectories obtained from the non adjusted fitted model does not produce constrained trajectories.
This suggests that we are indeed underestimating the smooth parameter $\omega$ with the tensor splines. This may be due to several factors : the low frequency at which the data is observed, the measurement error, the small amount of data that we have.
```{r}

cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N_ID+i)

  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,
                land=border,verbose=FALSE,save_omega = TRUE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore

  data_sim=cbind(data_sim,data_shore)
  data_sim
}

#stop cluster
stopCluster(cl)
```

@@

## Try a model directly on the whole data

@@

sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),length(all_data$time)),dim=c(2,2,length(all_data$time)))

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
                 nu=~ExpShip+s(ID,bs="re"),
                 omega=~te(AngleNormal,Ishore,k=5,bs="cs"))

response0<- SDE$new(formulas = formulas,data = all_data,type = "RACVM_SSM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))

response0$update_map(list("log_lambda"=factor(c(1,2,NA,NA))))
response0$update_lambda(exp(c(3,3,-15,-15)))

response0$fit()
```
```{r}
#plot parameters
xmin=list("AngleNormal"=-pi,"ExpShip"=1/45,"Ishore"=1/3)
xmax=list("AngleNormal"=pi,"ExpShip"=1/3,"Ishore"=1/0.1)
link=list("ExpShip"=(\(x) 1/x),"Ishore"=(\(x) 1/x))
xlabel=list("ExpShip"="Distance to ship","Ishore"="Distance to shore")
response0$get_all_plots(baseline=NULL,
      xmin=xmin, xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
```
